# Scenario II 


## Details

In this scenario a Gaussian prior on the effect size 
$\delta \sim \mathcal{N} ( 0.4, 0.2^2)$ is investigated. 
The null hypothesis is $\delta \leq 0$.
Currently, `adoptr` only supports normal distributed data what is widely
spread in the development of adaptive designs. 
We protect the one-sided type one error rate at $\alpha = 0.025$ and require
the power of the design to be at least $1 - \beta = 0.8$.



### Data distribution

Two-armed trial with normally distributed test statistic
```{r}
datadist <- Normal(two_armed = TRUE)
```


### Null hypothesis

The null hypothesis is $\mathcal{H}_0:\delta \leq 0$
```{r}
H_0 <- PointMassPrior(.0, 1)
```


### Prior assumptions
A Gaussian prior with mean $\delta = 0.4$ and standard deviation
$\tau = .2$ is defined.
```{r}
prior <- ContinuousPrior(function(delta) dnorm(delta, mean = .4, sd = .2),
                         support = c(-5, 5),
                         tighten_support = TRUE)
```




## Case II-1: Minimizing Expected Sample Size under Point Prior

### Objective

Expected sample size under the prior is minimized, i.e.,
$\boldsymbol{E}\big[n(\mathcal{D})\big]$.
```{r}
ess <- expected(ConditionalSampleSize(datadist, prior))
```


### Constrains

The type one error rate is controlled at $0.025$ on the boundary of the 
null hypothesis.
```{r}
toer_cnstr <- expected(ConditionalPower(datadist, H_0)) <= .025
```

Expected Power (rejection probability for positive effect sizes)
must be larger than $0.8$.
```{r}
pow_cnstr <- expected(
    ConditionalPower(datadist, condition(prior, c(0,3)))
    ) >= .8
```


### Initial Design

`adoptr` requires the definition of an initial design for optimization. 
We start with a group-sequential design from the package `rpact` that
fulfills the type-one error rate constraint and the power
constraint for a point effect size at $\delta = 0.4$.
The order of integration is set to $5$.
For usage as two-stage design with variable sample size, it has to
be converted to a `TwoStageDesign`.

```{r}
order <- 5L 

init_design_gs <- rpact_design(0.4, 0.025, 0.8, TRUE, order)

init_design    <- TwoStageDesign(init_design_gs)
```


### Optimization 

The optimal design is computed in three variants: two-stage,
group-sequentialand one-stage.
The input only differs with regard to the initial design.

```{r}
opt_design <- function(initial_design) {
    minimize(
        
        ess,
        
        subject_to(
            
            toer_cnstr,
            pow_cnstr
            
        ),
        
        initial_design = initial_design,
        
        opts = opts
        
)
}

opt1_gs <- opt_design(init_design_gs)
opt1_os <- opt_design(OneStageDesign(300, 2.0))
opt1_ts <- opt_design(TwoStageDesign(opt1_gs$design))
```



### Test Cases

Check if the optimization algorithm converged in all cases.
```{r}
iters <- sapply(list(opt1_ts, opt1_gs, opt1_os), 
                function(x) x$nloptr_return$iterations)

print(iters)

testthat::expect_true(all(iters < opts$maxeval))
```


Type one error rate constraint is tested for the three designs.
Due to numerical issues we allow a realtive error of $2\%$.
```{r}
tmp     <- sapply(list(opt1_ts, opt1_gs, opt1_os),  
                  function(x) ""sim_pr_reject(x$design, .0, datadist))
df_toer <- data.frame(
    toer = as.numeric(tmp[1, ]),
    se   = as.numeric(tmp[2, ])
)
rm(tmp)

testthat::expect_true(all(df_toer$toer <= .025*(1.02)))

df_toer
```



The expected sample sizes should be ordered in a specific way.
```{r}
testthat::expect_gte(
    evaluate(ess, opt1_os$design),
    evaluate(ess, opt1_gs$design)
)

testthat::expect_gte(
    evaluate(ess, opt1_gs$design),
    evaluate(ess, opt1_ts$design)
)
```





## Case II-2: Minimizing Expected Sample Size under Null Hypothesis

### Objective

Expected sample size conditioned on negative effect sizes is minimized, i.e.,
```{r}
ess_0 <- expected(ConditionalSampleSize(datadist, condition(prior, c(-3, 0))))
```


### Constrains

The constraints remain the same as before.

### Initial Design

The previous initial design can still be applied.


### Optimization 

The optimal group-sequential design and based on this the
optimal two-stage design are computed. 

```{r}
opt2 <- function(initial_design) {
    minimize(
        
        ess_0,
        
        subject_to(
            
            toer_cnstr,
            pow_cnstr
            
        ),
        
        initial_design = initial_design,
        
        opts = opts
        
)
}

opt2_gs <- opt2(init_design_gs)
opt2_ts <- opt2(TwoStageDesign(opt2_gs$design))
```



### Test Cases

Check if the optimization algorithm converged.
```{r}
print(opt2_ts$nloptr_return$iterations)

testthat::expect_true(opt2_ts$nloptr_return$iterations < opts$maxeval)
```


Type one error rate constraint is tested for the optimal design.
Due to numerical issues we allow a realtive error of $2\%$.
```{r}
tmp     <- ""sim_pr_reject(opt2_ts$design, .0, datadist)
df_toer2 <- data.frame(
    toer = as.numeric(tmp[1]),
    se   = as.numeric(tmp[2])
)
rm(tmp)

testthat::expect_true(all(df_toer2$toer <= .025*(1.02)))

df_toer2
```

The expected sample size under the null hypothesis should be lower
than of the design from variant II.1 where expected sample size under
the full prior was minimized.

```{r}
testthat::expect_lte(
    evaluate(ess_0, opt2_ts$design),
    evaluate(ess_0, opt1_ts$design)
)
```




## Case II-3: Conditional Power Constraint

### Objective

Expected sample size under the prior is minimized and has already been defined.

### Constrains

The constraints remain the same as before, additionally to a constraint
on conditional power.
```{r}
cp <- ConditionalPower(datadist, condition(prior, c(0, 3)))

cp_cnstr <- cp >= .7
```

### Initial Design

The previous initial design can still be applied.

### Optimization 

The optimal two-stage design is computed. 

```{r}
opt3_ts <- minimize(
        
        ess,
        
        subject_to(
            
            toer_cnstr,
            pow_cnstr,
            cp_cnstr
            
        ),
        
        initial_design = init_design,
        
        opts = opts
        
)
```



### Test Cases

Check if the optimization algorithm converged.
```{r}
print(opt3_ts$nloptr_return$iterations)

testthat::expect_true(opt3_ts$nloptr_return$iterations < opts$maxeval)
```


Type one error rate constraint is tested for the optimal design.
Due to numerical issues we allow a realtive error of $2\%$.
```{r}
tmp     <- ""sim_pr_reject(opt3_ts$design, .0, datadist)
df_toer3 <- data.frame(
    toer = as.numeric(tmp[1]),
    se   = as.numeric(tmp[2])
)
rm(tmp)

testthat::expect_true(all(df_toer3$toer <= .025*(1.02)))

df_toer3
```



The expected sample size under the prior should be higher than
in the case without the constraint that was analyzed in II.1.
```{r}
testthat::expect_gte(
    evaluate(ess, opt3_ts$design),
    evaluate(ess, opt1_ts$design)
)
```

The conditional power constraint needs to be tested. 
Select three points for this and check the constraint.
```{r}
x <- adoptr:::scaled_integration_pivots(opt3_ts$design)[c(1, 3, 5)]

cp_val <- sapply(x, function(z) evaluate(cp, opt3_ts$design, z))

testthat::expect_true(all(cp_val >= 0.7))
```



## Plot Two-Stage Designs
The optimal two-stage designs stemming from the different variants
are plotted together. 


```{r, echo = FALSE}
z1 <- seq(0, 4, by = .01)

tibble(
    type  = c("ESS under Prior", "ESS under Null", "ESS under Prior with CP constraint"), 
    design = list(opt1_ts$design, opt2_ts$design, opt3_ts$design)
) %>% 
    group_by(type) %>% 
    do(
        z1 = z1,
        n  = adoptr::n(.$design[[1]], z1),
        c2 = c2(.$design[[1]], z1),
        cp = evaluate(cp, .$design[[1]], z1)
    ) %>% 
    unnest() %>% 
    mutate(
        section = ifelse(
            is.finite(c2), 
            "continuation", 
            ifelse(c2 == -Inf, "efficacy", "futility")
        )
    ) %>% 
    gather(variable, value, n, c2, cp) %>% 
    ggplot(aes(z1, value, color = type)) +
        geom_line(aes(group = interaction(section, type))) + 
        facet_wrap(~variable, scales = "free_y") +
        theme_bw() +
        scale_color_manual(
          values = c(rgb(0,74,111, maxColorValue = 255),
                                    rgb(0,159,227, maxColorValue = 255),
                                    rgb(230,121,0, maxColorValue = 255))) +
        theme(
            panel.grid = element_blank(),
            legend.position = "bottom"
        )
```

