# Scenario IV: smaller effect, point prior {#scenarioIV}

## Details

In this scenario, we return to point priors as investigated in 
[Scenario I](#scenarioI). 
The main goal is to validate `adoptr`'s sensitivity with regard to 
the assumed effect size and the constraints on power and type one error rate.

Therefore, we still assume a two-armed trial with normally distributed outcomes.
The assumed effect size under the alternative is $\delta = 0.2$ in this setting. 
Type one error rate is protected at $2.5\%$ and the power should be at least
$80\%$. We will vary these values in the variants [IV.2](#variantIV.2) and
[IV.3](#variantIV.3)

```{r}
# data distribution and hypotheses
datadist   <- Normal(two_armed = TRUE)
H_0        <- PointMassPrior(.0, 1)
prior      <- PointMassPrior(.2, 1)

# constraints
alpha      <- 0.025
min_power  <- 0.8
toer_cnstr <- Power(datadist, H_0)   <= alpha
pow_cnstr  <- Power(datadist, prior) >= min_power
```



## Variant IV-1: Minimizing Expected Sample Size under Point Prior {#variantIV_1}

### Objective

Expected sample size under the alternative point prior $\delta = 0.2$
is minimized.

```{r objective}
ess <- ExpectedSampleSize(datadist, prior)
```


### Constraints
No additional constraints are considered in this variant.


### Initial Design

For this example, the optimal one-stage, group-sequential, and generic
two-stage designs are computed.
The initial design that is used as starting value of optimization is defined
as a group-sequential design by the package `rpact` that fulfills
type one error rate and power constraints in the case of group-sequential and
two-stage design. 
The initial one-stage design is chosen heuristically.
The order of integration is set to $5$.


```{r}
order <- 5L 

tbl_designs <- tibble(
    type    = c("one-stage", "group-sequential", "two-stage"),
    initial = list(
        OneStageDesign(500, 2.0),
        rpact_design(datadist, 0.2, 0.025, 0.8, TRUE, order),
        TwoStageDesign(rpact_design(datadist, 0.2, 0.025, 0.8, TRUE, order))) )
```


### Optimization 

```{r}
tbl_designs <- tbl_designs %>% 
    mutate(
       optimal = purrr::map(initial, ~minimize(
         
          ess,
          subject_to(
              toer_cnstr,
              pow_cnstr
          ),
          
          initial_design = ., 
          opts           = opts)) )
```



### Test Cases

Firstly, it is checked whether the maximum number of iterations was 
not exceeded in all three cases.

```{r}
tbl_designs %>% 
  transmute(
      type, 
      iterations = purrr::map_int(tbl_designs$optimal, 
                                  ~.$nloptr_return$iterations) ) %>%
  {print(.); .} %>% 
  {testthat::expect_true(all(.$iterations < opts$maxeval))}
```



Now, the constraints on type one error rate and power are tested via simulation.

```{r}
tbl_designs %>% 
  transmute(
      type, 
      toer  = purrr::map(tbl_designs$optimal, 
                         ~sim_pr_reject(.[[1]], .0, datadist)$prob), 
      power = purrr::map(tbl_designs$optimal, 
                         ~sim_pr_reject(.[[1]], .2, datadist)$prob) ) %>% 
  unnest(., cols = c(toer, power)) %>% 
  {print(.); .} %>% {
  testthat::expect_true(all(.$toer  <= alpha * (1 + tol)))
  testthat::expect_true(all(.$power >= min_power * (1 - tol))) }
```


Due to increasing degrees of freedom, the expected sample sizes under the
alternative should be ordered as 'one-stage > group-sequential > two-stage'.
They are evaluated by simulation as well as by `evaluate()`.

```{r}
tbl_designs %>% 
    mutate(
        ess      = map_dbl(optimal,
                           ~evaluate(ess, .$design) ),
        ess_sim  = map_dbl(optimal,
                           ~sim_n(.$design, .2, datadist)$n ) ) %>%
    {print(.); .} %>% {
    # sim/evaluate same under alternative?
    testthat::expect_equal(.$ess, .$ess_sim, 
                           tolerance = tol_n,
                           scale = 1)
    # monotonicity with respect to degrees of freedom
    testthat::expect_true(all(diff(.$ess) < 0)) }
```


Furthermore, the expected sample size under the alternative of the 
optimal group-sequential design should be lower than for the 
group-sequential design by `rpact` that is based on the inverse normal
combination test.

```{r}
tbl_designs %>%
             filter(type == "group-sequential") %>%
             { expect_lte(
                 evaluate(ess, {.[["optimal"]][[1]]$design}),
                 evaluate(ess, {.[["initial"]][[1]]})
             ) }
```

Finally, the $n_2$ function of the optimal two-stage design is expected to be 
monotonously decreasing:
```{r}
expect_true(
    all(diff(
        # get optimal two-stage design n2 pivots
        tbl_designs %>% filter(type == "two-stage") %>%
           {.[["optimal"]][[1]]$design@n2_pivots} 
        ) < 0) )
```




## Variant IV-2: Increase Power {#variantIV_2}

### Objective

The objective remains expected sample size under the alternative $\delta = 0.2$. 

### Constraints

The minimal required power is increased to $90\%$.

```{r}
min_power_2 <- 0.9
pow_cnstr_2 <- Power(datadist, prior) >= min_power_2
```


### Initial Design

For both flavours with two stages (group-sequential, generic two-stage)
the initial design is created by `rpact` to fulfill the error rate constraints.

```{r}
tbl_designs_9 <- tibble(
    type    = c("one-stage", "group-sequential", "two-stage"),
    initial = list(
        OneStageDesign(500, 2.0),
        rpact_design(datadist, 0.2, 0.025, 0.9, TRUE, order),
        TwoStageDesign(rpact_design(datadist, 0.2, 0.025, 0.9, TRUE, order))) )
```


### Optimization 

```{r}
tbl_designs_9 <- tbl_designs_9 %>% 
    mutate(
       optimal = purrr::map(initial, ~minimize(
         
          ess,
          subject_to(
              toer_cnstr,
              pow_cnstr_2
          ),
          
          initial_design = ., 
          opts           = opts)) )
```



### Test Cases

We start checking if the maximum number of iterations was not exceeded in all 
three cases.

```{r}
tbl_designs_9 %>% 
  transmute(
      type, 
      iterations = purrr::map_int(tbl_designs_9$optimal, 
                                  ~.$nloptr_return$iterations) ) %>%
  {print(.); .} %>% 
  {testthat::expect_true(all(.$iterations < opts$maxeval))}
```



The type one error rate and power constraints are evaluated by simulation.

```{r}
tbl_designs_9 %>% 
  transmute(
      type, 
      toer  = purrr::map(tbl_designs_9$optimal, 
                         ~sim_pr_reject(.[[1]], .0, datadist)$prob), 
      power = purrr::map(tbl_designs_9$optimal, 
                         ~sim_pr_reject(.[[1]], .2, datadist)$prob) ) %>% 
  unnest(., cols = c(toer, power)) %>% 
  {print(.); .} %>% {
  testthat::expect_true(all(.$toer  <= alpha * (1 + tol)))
  testthat::expect_true(all(.$power >= min_power_2 * (1 - tol))) }
```


Due to increasing degrees of freedom, the expected sample sizes under the
alternative should be ordered as 'one-stage > group-sequential > two-stage'.
This is tested by simulation as well as by `evaluate()`.

```{r}
tbl_designs_9 %>% 
    mutate(
        ess      = map_dbl(optimal,
                           ~evaluate(ess, .$design) ),
        ess_sim  = map_dbl(optimal,
                           ~sim_n(.$design, .2, datadist)$n ) ) %>%
    {print(.); .} %>% {
    # sim/evaluate same under alternative?
    testthat::expect_equal(.$ess, .$ess_sim, 
                           tolerance = tol_n,
                           scale = 1)
    # monotonicity with respect to degrees of freedom
    testthat::expect_true(all(diff(.$ess)     < 0)) 
    testthat::expect_true(all(diff(.$ess_sim) < 0))}
```

Comparing with the inverse-normal based group-sequential design created
by `rpact`, the optimal group-sequential design should show
a lower expected sample size under the point alternative.

```{r}
tbl_designs_9 %>%
             filter(type == "group-sequential") %>%
             { expect_lte(
                 evaluate(ess, {.[["optimal"]][[1]]$design}),
                 evaluate(ess, {.[["initial"]][[1]]})
             ) }
```

Since a point prior is regarded, the $n_2$ function of the optimal 
two-stage design is expected to be monotonously decreasing:
```{r}
expect_true(
    all(diff(
        # get optimal two-stage design n2 pivots
        tbl_designs_9 %>% filter(type == "two-stage") %>%
           {.[["optimal"]][[1]]$design@n2_pivots} 
        ) < 0) )
```



## Variant IV-3: Increase Type One Error rate {#variantIV_3}

### Objective

As in variants [IV.1](#variantIV_1) and [IV-2](#variantIV_2),
expected sample size under the point alternative is minimized.

### Constraints

While the power is still lower bounded by $90\%$ as in variant [II](#variantIV_2),
the maximal type one error rate is increased to $5\%$.

```{r}
alpha_2      <- .05
toer_cnstr_2 <- Power(datadist, H_0) <= alpha_2
```

### Initial Design

Again, a design computed by means of the package `rpact` to fulfill
the updated error rate constraints is applied as initial design for the 
optimal group-sequential and generic two-stage designs.

```{r}
tbl_designs_5 <- tibble(
    type    = c("one-stage", "group-sequential", "two-stage"),
    initial = list(
        OneStageDesign(500, 2.0),
        rpact_design(datadist, 0.2, 0.05, 0.9, TRUE, order),
        TwoStageDesign(rpact_design(datadist, 0.2, 0.05, 0.9, TRUE, order))) )
```


### Optimization 

```{r}
tbl_designs_5 <- tbl_designs_5 %>% 
    mutate(
       optimal = purrr::map(initial, ~minimize(
         
          ess,
          subject_to(
              toer_cnstr_2,
              pow_cnstr_2
          ),
          
          initial_design = ., 
          opts           = opts)) )
```



### Test Cases

The convergence of the optimization algorithm is tested by checking if the
maximum number of iterations was not exceeded.

```{r}
tbl_designs_5 %>% 
  transmute(
      type, 
      iterations = purrr::map_int(tbl_designs_5$optimal, 
                                  ~.$nloptr_return$iterations) ) %>%
  {print(.); .} %>% 
  {testthat::expect_true(all(.$iterations < opts$maxeval))}
```


By simulation, the constraints on the error rates (type one error and power)
are tested.

```{r}
tbl_designs_5 %>% 
  transmute(
      type, 
      toer  = purrr::map(tbl_designs_5$optimal, 
                         ~sim_pr_reject(.[[1]], .0, datadist)$prob), 
      power = purrr::map(tbl_designs_5$optimal, 
                         ~sim_pr_reject(.[[1]], .2, datadist)$prob) ) %>% 
  unnest(., cols = c(toer, power)) %>% 
  {print(.); .} %>% {
  testthat::expect_true(all(.$toer  <= alpha_2 * (1 + tol)))
  testthat::expect_true(all(.$power >= min_power_2 * (1 - tol))) }
```


Due to increasing degrees of freedom, the expected sample sizes under the
alternative should be ordered as 'one-stage > group-sequential > two-stage'.
They are tested by simulation as well as by calling `evaluate()`.

```{r}
tbl_designs_5 %>% 
    mutate(
        ess      = map_dbl(optimal,
                           ~evaluate(ess, .$design) ),
        ess_sim  = map_dbl(optimal,
                           ~sim_n(.$design, .2, datadist)$n ) ) %>%
    {print(.); .} %>% {
    # sim/evaluate same under alternative?
    testthat::expect_equal(.$ess, .$ess_sim, 
                           tolerance = tol_n,
                           scale = 1)
    # monotonicity with respect to degrees of freedom
    testthat::expect_true(all(diff(.$ess) < 0)) }
```


The expected sample size under the alternative that was used as objective criterion
of the optimal group-sequential design should be lower than for the 
group-sequential design by `rpact` that is based on the inverse normal
combination test.

```{r}
tbl_designs_5 %>%
             filter(type == "group-sequential") %>%
             { expect_lte(
                 evaluate(ess, {.[["optimal"]][[1]]$design}),
                 evaluate(ess, {.[["initial"]][[1]]})
             ) }
```

Also in this variant, the $n_2$ function of the optimal two-stage design 
is expected to be monotonously decreasing:
```{r}
expect_true(
    all(diff(
        # get optimal two-stage design n2 pivots
        tbl_designs_5 %>% filter(type == "two-stage") %>%
           {.[["optimal"]][[1]]$design@n2_pivots} 
        ) < 0) )
```




## Plot Two-Stage Designs
The optimal two-stage designs stemming from the three different variants
are plotted together. 

```{r, echo=FALSE}
x1 <- seq(-.5, 3, by = .01)

tibble(
    constraints  = c(
        "TOER<=0.025, Power>=0.8",
        "TOER<=0.025, Power>=0.9",
        "TOER<=0.050, Power>=0.9" ), 
    design = list(
        tbl_designs %>% 
            filter(type == "two-stage") %>% 
            pull(optimal) %>% 
            .[[1]] %>% 
            .$design, 
        tbl_designs_9 %>% 
            filter(type == "two-stage") %>% 
            pull(optimal) %>% 
            .[[1]] %>% 
            .$design, 
        tbl_designs_5 %>% 
            filter(type == "two-stage") %>% 
            pull(optimal) %>% 
            .[[1]] %>% 
            .$design ) ) %>% 
    group_by(constraints) %>% 
    do(
        x1 = x1,
        n  = adoptr::n(.$design[[1]], x1),
        c2 = c2(.$design[[1]], x1) ) %>% 
    unnest(., cols = c(x1, n, c2)) %>% 
    mutate(
        section = ifelse(
            is.finite(c2), 
            "continuation", 
            ifelse(c2 == -Inf, "efficacy", "futility") ) ) %>% 
    gather(variable, value, n, c2) %>% 
    ggplot(aes(x1, value, color = constraints)) +
        geom_line(aes(group = interaction(section, constraints))) + 
        facet_wrap(~variable, scales = "free_y") +
        labs(y = "", x = expression(x[1])) +
        scale_color_discrete("") +
        theme_bw() +
        theme(
            panel.grid      = element_blank(),
            legend.position = "bottom" )
```

