# Scenario VI: binomial distribution {#scenarioVI}

## Details

In this scenario, the implemented normal approximation of the binomial
distribution is evaluated.
There are two treatment groups $E$ (experimental) and $C$ (control)
and the outcome of interest is the event rate in both groups.
The event rate in the control group is assumed to equal $p_C = 0.3$.
Under the null hypothesis, the rate difference $p_E - p_C = 0$ is assumed.
Under the alternative, the rate difference is assumed to equal $p_E - p_C = 0.2$.
Type one error rate should be protected at 2.5\% and the design's power should
be at least 90\%.

```{r}
# data distribution and hypotheses
datadist   <- Binomial(rate_control = 0.3, two_armed = TRUE)
H_0        <- PointMassPrior(.0, 1)
prior      <- PointMassPrior(.2, 1)

# define constraints
alpha      <- 0.025
min_power  <- 0.9
toer_cnstr <- Power(datadist, H_0)   <= alpha
pow_cnstr  <- Power(datadist, prior) >= min_power
```



## Variant VI-1, efficiency of optimal designs for binomial distribution {#variantVI_1}

In this variant, it is evaluated whether the optimization leads to more
efficient designs and is plausible with published results.

### Objective

Expected sample size under the alternative point mass prior $\delta = 0.2$
is minimized.
```{r}
ess <- ExpectedSampleSize(datadist, prior)
```


### Constraints

No additional constraints are considered.


### Initial Designs

For this example, the optimal one-stage, group-sequential, and generic
two-stage designs are computed.
The initial design for the one-stage case is determined heuristically
and both the group sequential and the generic two-stage designs are 
optimized starting from the corresponding group-sequential design as
computed by the `rpact` package.
```{r}
order <- 7L
# data frame of initial designs 
tbl_designs <- tibble(
    type    = c("one-stage", "group-sequential", "two-stage"),
    initial = list(
        OneStageDesign(200, 2.0),
        rpact_design(datadist, 0.2, 0.025, 0.9, TRUE, order),
        TwoStageDesign(rpact_design(datadist, 0.2, 0.025, 0.9, TRUE, order))) )
```

The order of integration is set to `r order`.


### Optimization 

```{r}
tbl_designs <- tbl_designs %>% 
    mutate(
       optimal = purrr::map(initial, ~minimize(
         
          ess,
          subject_to(
              toer_cnstr,
              pow_cnstr
          ),
          
          initial_design = ., 
          opts           = opts)) )
```



### Test Cases

To avoid improper solutions, it is first verified that the maximum
number of iterations was not exceeded in any of the three cases.
```{r}
tbl_designs %>% 
  transmute(
      type, 
      iterations = purrr::map_int(tbl_designs$optimal, 
                                  ~.$nloptr_return$iterations) ) %>%
  {print(.); .} %>% 
  {testthat::expect_true(all(.$iterations < opts$maxeval))}
```


Next, the type one error rate and power constraints are verified 
for all three designs by simulation:
```{r}
tbl_designs %>% 
  transmute(
      type, 
      toer  = purrr::map(tbl_designs$optimal, 
                         ~sim_pr_reject(.[[1]], .0, datadist)$prob), 
      power = purrr::map(tbl_designs$optimal, 
                         ~sim_pr_reject(.[[1]], .2, datadist)$prob) ) %>% 
  unnest(., cols = c(toer, power)) %>% 
  {print(.); .} %>% {
  testthat::expect_true(all(.$toer  <= alpha * (1 + tol)))
  testthat::expect_true(all(.$power >= min_power * (1 - tol))) }
```



Since the degrees of freedom of the three design classes are ordered as
'two-stage' > 'group-sequential' > 'one-stage',
the expected sample sizes (under the alternative) should be ordered 
in reverse ('two-stage' smallest).
Additionally, expected sample sizes under both null and alternative
are computed both via `evaluate()` and simulation-based.
```{r}
ess0 <- ExpectedSampleSize(datadist, H_0)
tbl_designs %>% 
    mutate(
        ess      = map_dbl(optimal,
                           ~evaluate(ess, .$design) ),
        ess_sim  = map_dbl(optimal,
                           ~sim_n(.$design, .2, datadist)$n ),
        ess0     = map_dbl(optimal,
                           ~evaluate(ess0, .$design) ),
        ess0_sim = map_dbl(optimal,
                           ~sim_n(.$design, .0, datadist)$n ) ) %>% 
    {print(.); .} %>% {
    # sim/evaluate same under alternative?
    testthat::expect_equal(.$ess, .$ess_sim, 
                           tolerance = tol_n,
                           scale = 1)
    # sim/evaluate same under null?
    testthat::expect_equal(.$ess0, .$ess0_sim, 
                           tolerance = tol_n,
                           scale = 1)
    # monotonicity with respect to degrees of freedom
    testthat::expect_true(all(diff(.$ess) < 0)) }
```

The expected sample size under the alternative must be lower or equal than
the expected sample size of the initial `rpact` group-sequential design that
is based on the inverse normal combination test.
```{r}
testthat::expect_lte(
  evaluate(ess, 
             tbl_designs %>% 
                filter(type == "group-sequential") %>% 
                pull(optimal) %>% 
                .[[1]]  %>%
                .$design ),
    evaluate(ess, 
             tbl_designs %>% 
                filter(type == "group-sequential") %>% 
                pull(initial) %>% 
                .[[1]] ) )
```


Finally, the `rpact` group-sequential design,
the optimal group-sequential design, and the optimal generic two-stage design
are plotted to allow visual inspection.

```{r, echo = FALSE}
x1 <- seq(-.25, 2.5, by = .01)

tibble(
    type  = c(
        "rpact", 
        "optimal group-sequential", 
        "optimal two-stage" ), 
    design = list(
        tbl_designs %>% 
                filter(type == "group-sequential") %>% 
                pull(initial) %>% 
                .[[1]] , 
        tbl_designs %>% 
                filter(type == "group-sequential") %>% 
                pull(optimal) %>% 
                .[[1]] %>%
                .$design, 
        tbl_designs %>% 
                filter(type == "two-stage") %>% 
                pull(optimal)  %>% 
                .[[1]] %>%
                .$design ) ) %>% 
  # plot the designs
    group_by(type) %>% 
    do(
        x1 = x1,
        n  = adoptr::n(.$design[[1]], x1),
        c2 = c2(.$design[[1]], x1),
        CP = evaluate(ConditionalPower(datadist, prior), .$design[[1]], x1) ) %>% 
    unnest(., cols = c(x1, n, c2, CP)) %>% 
    mutate(
        section = ifelse(
            is.finite(c2), 
            "continuation", 
            ifelse(c2 == -Inf, "efficacy", "futility") ) ) %>% 
    gather(variable, value, n, c2, CP) %>% 
    ggplot(aes(x1, value, color = type)) +
        geom_line(aes(group = interaction(section, type))) + 
        facet_wrap(~variable, scales = "free_y") +
        labs(y = "", x = expression(x[1])) +
        scale_color_discrete("") +
        theme_bw() +
        theme(
            panel.grid       = element_blank(),
            legend.direction = "vertical",
            legend.position  = "bottom" )
```
